---
title: "NOAA Storm Analysis"
author: "Nitesh"
date: "3 September 2016"
output: 
  html_document:
    keep_md: true
---

## 1.Synopsis
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. At first the data is downloaded if it is already not present and is then loaded into the workspace. Then the data is analysed for further study. Here two questions have been answered, which types of events are most harmful with respect to population health and which types of events have the greatest economic consequences.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
```

### Loading Required Packages
```{r}
library(ggplot2)
library(plyr)
library(grid)
library(gridExtra)
```




## 2.Data Processing

### 2.1.Downloading Data
```{r}
if(!file.exists("repdata%2Fdata%2FStormData.csv.bz2"))
{
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url,"repdata%2Fdata%2FStormData.csv.bz2")
}
```

### 2.2.Loading Data
```{r}
StormData <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
```
### 2.3.Reducing The Dataset
Since the latest years account for most of the data, initial values(25%) are being been removed.

```{r}
ReducedRows <- 0.25 * nrow(StormData)
StormData <- StormData[ReducedRows:nrow(StormData),]
```

Some columns contain NA values and need to be removed, thus removing those columns

```{r}
StormData <- StormData[,colSums(is.na(StormData))==0]
```

### 2.4.Refactoring EVTYPE With Similar Names

Determing all the event type

```{r}
head(unique(StormData$EVTYPE),n = 50)
```
It can be seen that there are event out of which many are of similar type and should have fallen under the same category

Hence, renaming event type :

```{r}
##Converting EVTYPE from factor to character
StormData$EVTYPE <- as.character(StormData$EVTYPE)
StormData[grepl("tornado|gustnado|spout|funnel|whirlwind",StormData$EVTYPE,ignore.case=TRUE),"EVTYPE"] <- "Tornado"
StormData[grepl("wind|storm|wnd|hurricane|typhoon",StormData$EVTYPE,ignore.case=TRUE),"EVTYPE"] <- "Wind"
StormData[grepl("tstm|thunderstorm|lightning|lighting",StormData$EVTYPE,ignore.case=TRUE),"EVTYPE"] <- "Lightning & Thunderstorm"
StormData[grepl("flood|rising water|floooding|surf|blow-out|swells|fld|dam break|seas|high water|tide|tsunami|wave|current|marine|drowning|stream|beach erosin|seiche" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Flood & Tsunami"
StormData[grepl("dust|saharan" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Dust"
StormData[grepl("fire|smoke|volcanic" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Fire & Volcano"
StormData[grepl("cold|cool|ice|icy|frost|freeze|snow|winter|wintry|wintery|blizzard|chill|freezing|avalanche|glaze|sleet|low temperature" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Cold & Snow"
StormData[grepl("slide|erosion|slump" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Landslide"
StormData[grepl("warmth|warm|heat|dry|hot|drought|driest|thermia|temperature record|record temperature|record high" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Heat & Drought"
StormData[grepl("precipitation|shower|rain|hail|drizzle|wet|percip|burst|depression|fog|wall cloud|vog|precip" , StormData$EVTYPE, ignore.case=TRUE), "EVTYPE"] <- "Rainfall & Fog"
```
Removing the events whose type is not known

```{r}
StormData <- StormData[!grepl("summary|none|other|excessive|floyd|criteria|monthly temperature|mild pattern|no severe weather|southeast|urban small|heavy mix|northern lights|\\?",StormData$EVTYPE,ignore.case = TRUE),]
```

Plot showing count of events occuring

```{r}
ggplot(data = StormData,aes(EVTYPE,fill = EVTYPE)) + geom_bar() + ggtitle("Count Of Each Event") + xlab("Event Type") + ylab("Count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### 2.5.Assigning Proper Values To Required Columns

```{r}
TenthPower <- function(x)
{
    if(is.numeric(x)) {
      x <- x
    }
    else if(grepl("h", x, ignore.case=TRUE)) {
      x <- 2
    }
    else if(grepl("k", x, ignore.case=TRUE)) {
      x <- 3
    }
    else if(grepl("m", x, ignore.case=TRUE)) {
      x <- 6
    }
    else if(grepl("b", x, ignore.case=TRUE)) {
      x <- 9
    }
    else if(x == "" | x == " "){
      x <- 0
    }
    else{
      x <- NA
    }
    return(x)
}
calculateDamage <- function(num, exponent)
{
    pow <- TenthPower(exponent)
    if(is.numeric(num))
    {
      num <- num * (10 ^ pow)
    }
    if(!is.numeric(num))
    {
      num <- 0
    }
    num
}

StormData$propDamage <- mapply(calculateDamage, StormData$PROPDMG, StormData$PROPDMGEXP)
StormData$cropDamage <- mapply(calculateDamage, StormData$CROPDMG, StormData$CROPDMGEXP)
```

### 2.6.Types of events most harmful to population health

```{r}
StormData$HealthDamage <- StormData$FATALITIES + StormData$INJURIES
PopulationDamage <- aggregate(HealthDamage ~ EVTYPE,data = StormData,sum)
```



### 2.7.Types of events having the greatest economic consequences

```{r}
StormData$TotalDamage = StormData$propDamage + StormData$cropDamage
EconomicDamage <- aggregate(TotalDamage ~ EVTYPE,data = StormData,sum)
```

## 3.Results

### 3.1.Tables
Most harmful events to human health arranged in the order to high to low

```{r}
PopulationDamage[with(PopulationDamage,order(-HealthDamage)),]
```

Events with the greatest economic consequences arranged in the order high to low

```{r}
EconomicDamage[with(EconomicDamage,order(-TotalDamage)),]
```

### 3.2.Figures
Plot to show that damage to health by the events is different from economic consequences

```{r}
plot1 <- ggplot(data = EconomicDamage,aes(EVTYPE,TotalDamage)) + geom_point() + ggtitle("Economic Consequnces") + ylab("Total Damage(million $)") +theme(axis.text.x = element_text(angle = 90, hjust = 1))
plot2 <- ggplot(data = PopulationDamage,aes(EVTYPE,HealthDamage)) + geom_point() + ggtitle("Health Casualities") + ylab("Health Damage") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
grid.arrange(plot1,plot2,ncol = 2,top=textGrob("Comparison", gp = gpar(fontsize=20,font=8)))
```









